Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-26 article.tex
Retouches Sections 1+2

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows: Section
\ref{sec:filter} describes the signal processing methods applied on fluorescent
calcium time series; Section \ref{sec:inference} then presents the proposed
approach and its theoretical properties; Section \ref{sec:results} provides an
empirical analysis and comparison with other network inference methods, while
Section \ref{sec:conclusion} finally discusses our work and provides further
research directions. Additionally, Appendix \ref{app:optimized} further
describes, in all details, our actual winning method, giving slightly better
results than the method presented in the paper at a price of significant
tuning.

@@

% \begin{figure}
% \centering
% \includegraphics[width=0.4\linewidth]{images/fig_filtering}
% \caption{Signal processing process for extracting peaks from the raw fluorescence data.}
% \label{fig:filtered-signal}
% \end{figure}

@@

Furthermore, it can be observed that neurons communicate through short spikes, characterized by a high
frequency, while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i \label{eq:high-pass-filter},
\end{align}
shown in the plot of Figure \ref{fig:hp_curve}:

To filter out small variations in the signal obtained after applying the function $g$ as well as to eliminate negative values, we use the following hard-threshold filter
\begin{align}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding the Figure \ref{fig:threshold_curve}.
As we can see, the processed signal only contains clean spikes.

The objective of the last step of our filtering procedure is to decrease the importance of spikes that occur when there is a high global activity in the network with respect to  spikes that occur during normal activity. Indeed, we have conjectured that when a large part of the network is firing, the rate at which observations are made is not high enough for being able to detect interactions and, that it would therefore be preferrable to lower their importance by changing appropriatly their magnitude. Additionnally, it is well-known that neurons may also spike because of a global high activity (see \cite{stetter2012model}). In such  context, detecting pairwise neuron interactions from the firing activity is meaningless. 
Accordingly, we have chosen to apply to the signal outputted by $h$ the following function  
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}.
\end{align}
which effect is to magnify the importance of spikes that occur in case of low global activity. Note that the particular case where $\sum_{j} x^{t}_j$ equals $0$ is solved by setting $w(x^{t}_i) = 1$.

>>>>

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows: Section
\ref{sec:filter} describes the signal processing methods applied on fluorescent
calcium time series; Section \ref{sec:inference} then presents the proposed
approach and its theoretical properties; Section \ref{sec:results} provides an
empirical analysis and comparison with other network inference methods, while
Section \ref{sec:conclusion} finally discusses our work and provides further
research directions. Additionally, Appendix \ref{app:optimized} further
describes, in all details, our actual winning method, giving slightly better
results than the method presented in the paper, at the price of parameter
tuning.

@@

@@

Furthermore, it can be observed that neurons communicate through short spikes, characterized by a high
frequency, while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i \label{eq:high-pass-filter},
\end{align}
as shown in Figure \ref{fig:hp_curve}.

To filter out small variations in the signal obtained after applying the
function $g$ as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding Figure \ref{fig:threshold_curve}.
As we can see, the processed signal only contains clean spikes.

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is a high global activity in the
network with respect to  spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough for being able to detect
interactions and, that it would therefore be preferrable to lower their
importance by changing appropriatly their magnitude. Additionnally, it is
well-known that neurons may also spike because of a global high activity (see
\cite{stetter2012model}). In such  context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such,
the signal output by $h$ is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}},
\end{align}
which effect is to magnify the importance of spikes that occur in case of low
global activity. Note that the particular case where $\sum_{j} x^{t}_j = 0$ is
solved by setting $w(x^{t}_i) = 1$.
